### 2022-05-30 src/ms.tex
Finish writing the microlensing part of Chapter 2 (for now)

% Define document class
\documentclass[11pt]{report}
\usepackage{aas_macros}

@@

That the same effect could be used to detect the presence of planets orbiting
around the lens star was first theorized by \citet{1964PhRv..133..835L} who
wrote that \emph{``the primary effect of planetary deflectors bound to stars
    other than the Sun will be to slightly perturb the lens action of these
    stars''} although he was also sceptical about the possibility of detection,
saying that \emph{``associated pulses would be so weak and infrequent and of
    such fleeting duration – perhaps a few hours – as to defy detection''}.
Gravitational lensing as a method for discovering exoplanets really took off
with the work of Paczyński
\citep{1986ApJ...304....1P,1986ApJ...301..503P,1991ApJ...374L..37M} who also
coined the term \emph{microlensing}, which refers to gravitational lensing in a
regime where the images of the background object cannot be resolved but one can
nevertheless measure its magnification as a function of time. Microlensing as a
method for detecting exoplanets has some unique aspects. First, it is a one-off
event which happens on a timescale of a few minutes up to several months
depending on the distances to the background star and the lensing star and the
mass of the lens. In addition to the fact that there's only one chance to
observe such an event for it to happen at all requires extremely precise
alignment between the lens and the background star and the chance of observing
a stellar microlensing event is about one-in-a-million (CITE) for a typical
star within the Milky Way. To observe a planetary signal is about an order of
magnitude less likely than that. Hence, obtaining a decent sample of planetary
events requires continuous monitoring on the order of $10^{8}$ stars at the
time which way microlensing observations that to focus on the densest region of
the Milky Way -- the galactic bulge. Figure~\ref{fig:crowded_field} a picture
from of such a dense stellar field. Finally, in the vast majority of cases we
do not detect any light from the lens itself, the collected photons are from a
background star completely unrelated and distant from the lensing star. This is
very different from other exoplanet discovery methods and it means that we only
obtain dynamical properties of planets such their masses and periods.

@@

From Figure~\ref{fig:lens_geometry} it follows that the relationship between
the observed and the actual location of the source is given by
\begin{equation}
    \beta=\theta-2 R_s \frac{D_S-D_L}{D_LD_S}
    \frac{1}{\theta}\quad,
    \label{eq:lens_equation}
\end{equation}
which is known as the \emph{lens equation} or sometimes also the \emph{ray-tracing}
equation. If the source and the lens are perfectly aligned with respect to the observer,
then $\beta=0$ and we have
\begin{equation}
    \theta \equiv \theta_E= \sqrt{ \frac{4GM}{c^2} \left( \frac{1}{D_L} - \frac{1}{D_S} \right)}
\end{equation}
The distance term $D_L^{-1}-D_\textrm{s}^{-1}$ is proportional to the
lens-source parallax.

@@

%The angular Einstein radius determines a characteristic timescale
%$t_E=\theta_E/|\boldsymbol{\mu}_{LS}|$ on which the magnification of of the
%source changes, where $\boldsymbol{\mu}_{LS}$ is the relative proper motion
%vector between the source and the lens
%$\boldsymbol{\mu}_{LS}\equiv\boldsymbol{\mu}_L- \boldsymbol{\mu}_S$. The
%predicted flux of a star undergoing a microlensing event is then given by the
%following expression
%\begin{equation}
%    f=F_SA(t)+F_B
%    \label{eq:flux_paczynski}
%\end{equation}
%where $F_S$ is the intrinsic flux of the star and
%$F_B$ is any background flux due to other stars in the field which
%might be blended with the source star, usually called the \emph{blending flux}.
%$A(t)$ is a non-linear time-dependant magnification mapping which depends
%on the properties of the lens and the geometry of the lensing event.
%In the simplest case, that of a point source being magnified by a point lens,
%the magnification is given by
%\begin{equation}
%    A(t)= \frac{u(t)^2+2}{u(t)\sqrt{u(t)^2+4}}
%\end{equation}
%where $u(t)$ is the separation between the lens and the source expressed in
%units of the angular Einstein radius.

To be able to evaluate Equation~\ref{eq:magnification_point_lens} in practice
we have to parametrize the position of the source on the sky $u$ as a function
of time. To derive an expression for $u(t)$ we assume the the motion of the
observer, lens and the source is rectilinear (acceleration can be neglected).
We take $\boldsymbol{\theta}_S$ to be the angular position of the source star
on the plane of the sky and $\boldsymbol{\mu}_L$ to be its proper motion
vector. Likewise for the lens. We thus have
\begin{align}
    \boldsymbol{\theta}_S(t) & =\boldsymbol{\theta}_{S, 0}+
    \left(t-t_{0}\right) \boldsymbol{\mu}_S
    \label{eq:source_position}                              \\
    \boldsymbol{\theta}_L(t) & =\boldsymbol{\theta}_{L, 0}
    +\left(t-t_{0}\right) \boldsymbol{\mu}_L\quad,
    \label{eq:lens_position}
\end{align}
where $t_0$ is some reference time.
The relative position vector of the lens with respect to the source is then
\begin{equation}
    \boldsymbol{u}(t) \equiv \frac{\boldsymbol{\theta}_L(t)
        -\boldsymbol{\theta}_S(t)}{\theta_E}=
    \frac{\boldsymbol{\theta}_{LS, 0}}{\theta_E}
    +\frac{t-t_{0}}{\theta_E} \boldsymbol{\mu}_{LS}
    \label{eq:relative_trajectory_no_parallax}
\end{equation}
where
$\boldsymbol{\theta}_{L S, 0}\equiv\boldsymbol{\theta}_{L,0}-\boldsymbol{\theta}_{S, 0}$
is the relative position at $t_0$ and
$\boldsymbol{\mu}_{LS}\equiv \boldsymbol{\mu}_{L}- \boldsymbol{\mu}_{S}$ is the
relative proper motion.
Since $\boldsymbol{\mu}_{LS}$ and $\boldsymbol{\theta}_{L S, 0}$ are
perpendicular to each other, it follows that the magnitude of the relative
separation is
\begin{equation}
    u(t)=\sqrt{u_0^2+ \left(\frac{t-t_0}{t_E}\right)^2}
\end{equation}
where we have defined $u_0\equiv |\boldsymbol{\theta}_{L S, 0}|/\theta_E$
and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
The model for the light curve is then characterized by
the parameters $\{F_S,F_B, t_0, u_0, t_E\}$ and is known in the
literature as the Paczy\'nski curve.

The magnification as a function of time for different impact parameters $u_0$
is shown in Figure~\ref{fig:single_lens_mag}. Notice the very steep fall off in
magnification as as we move away from the Einstein ring ($A(u)\propto
    1+2u^{-4}$). The curves shown in Figure~\ref{fig:paczynski_curve} are often
called \emph{Paczyński curves} \citep{1986ApJ...304....1P,1986ApJ...301..503P},
in honour of the late Polish astronomer Bohdan Paczyński, one of the key
figures of gravitational microlensing.

@@

@@

\subsection{Parallax}
In Equation~\ref{eq:relative_trajectory_no_parallax} we have assumed that the
relative motion of the lens with respect to the source on the plane of the sky
is rectilinear. This is a reasonable approximation in a barycentric frame of
reference if the acceleration of the source and the lens can be neglected.
However, most microlensing events are observed from a non-inertial geocentric
frame so that the observed trajectory is modified due the local acceleration of
the Earth and this effect needs to be taken into account in a realistic model.
In order to include the parallax effect, we need to modify
Equations~\ref{eq:source_position} and \ref{eq:lens_position} to include the
projected motion of the Sun relative to the Earth on the plane of the sky. We
project $\mathbf s$ -- the position vector of the Sun relative to Earth, onto a
plane perpendicular to the line of sight towards the source star (plane of the
sky) which is defined by the unit vector $\boldsymbol{\hat n}$ normal to the
plane. The unit vector $\boldsymbol{\hat n}$ depends on the sky coordinates of
the source star $(\alpha,\delta)$ (right ascension and declination).

We choose to work in geocentric equatorial coordinates defined by the spherical
unit vectors $\hat{\mathbf e}_n$, pointing North, and $\hat{\mathbf e}_e$
pointing eastward such that the coordinate system is right-handed. The unit
vectors are defined by
\begin{align}
    \hat{\mathbf e}_e & = \hat{\mathbf z}\times \hat{\mathbf n}  \\
    \hat{\mathbf e}_n & = \hat{\mathbf n}\times\hat{\mathbf e}_e
\end{align}
The two components of $\mathbf s$ projected onto the plane of the sky are then
\begin{align}
    \zeta_e(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_e \\
    \zeta_n(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_n
\end{align}
In practice, we can retrieve $\mathbf{s}$ using NASA's JPL Horizons system and compute
the projected separation at any time $t$.
The angular positions of the source and the lens
(Equations~\ref{eq:source_position} and \ref{eq:lens_position}) then become
\begin{align}
    \boldsymbol{\theta}_S(t) & = \boldsymbol{\theta}_{S,0}+(t-t_0)\boldsymbol{\mu}_S
    +\pi_S\,\boldsymbol{\zeta}(t)                                                    \\
    \boldsymbol{\theta}_L(t) & = \boldsymbol{\theta}_{L,0}+(t-t_0)\boldsymbol{\mu}_L
    +\pi_L\,\boldsymbol{\zeta}(t)
\end{align}
where  $\pi_S\equiv 1\,\mathrm{au}/D_S$ is the source parallax, and $\pi_L$ is the
lens parallax.
The relative separation is then
\begin{equation}
    \boldsymbol{u}(t)= \frac{\boldsymbol\theta_{LS, 0}}{\theta_E}
    +\frac{t-t_0}{\theta_E}\boldsymbol{\mu}_{LS}+\pi_{E}\,\boldsymbol{\zeta}(t)
    \label{eq:relative_separation_parallax}
\end{equation}
with $\pi_E\equiv\pi_{LS}/\theta_E\equiv (\pi_L-\pi_S)/\theta_E$.

Since annual parallax is a higher order effect affecting the apparent
trajectory of the lens on the sky, it makes sense to decompose the trajectory
as a sum of rectilinear motion plus a deviation due to parallax. Following
\citet{2004ApJ...606..319G} we define the position offset of the Sun on the
plane of the sky relative to its position at a particular time $t_0'$ as

@@

\begin{figure}
    \centering
    \includegraphics[width=0.3\linewidth]{static/microlensing/microlensing_parallax_coordinates.pdf}
    \caption{A coordinate system
        $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ parallel to the lens trajectory at time $t_0$. The orange curve
        represents the lens trajectory relative to the source star at the origin. The
        coordinate system $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ is related to the equatorial coordinates $(\mathbf{\hat
                e}_e,\mathbf{\hat e}_n)$ by a rotation through an angle $\psi$.}
    \label{fig:parallax2}
\end{figure}

To evaluate Equation~\ref{eq:relative_separation_parallax_decomposed}, we need
to choose a suitable basis. A natural coordinate system for describing the
trajectory of the lens relative to the source is one defined by unit vectors
$(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ where $\mathbf{\hat e}_\parallel$ is parallel to the trajectory $\mathbf{u}(t)$ at time $t_0$
(Figure~\ref{fig:parallax2}). We define the unit vectors as
\begin{equation}
    \mathbf{\hat e}_\bot\equiv \frac{\mathbf{u}_0}{|\mathbf{u}_0|}\quad
    \mathbf{\hat e}_\parallel\equiv \frac{\mathbf{\hat n}\times\mathbf{u}_0}{|\mathbf{u}_0|}\quad
\end{equation}
and the coordinate system $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ is related to ecliptic coordinates via a simple rotation through an
angle $\psi$
\begin{equation}
    \begin{pmatrix}
        \mathbf{\hat e}_\bot \\
        \mathbf{\hat e}_\parallel
    \end{pmatrix}
    =
    \begin{pmatrix}
        \cos\psi & -\sin\psi \\
        \sin\psi & \cos\psi
    \end{pmatrix}
    \begin{pmatrix}
        \mathbf{\hat e}_e \\
        \mathbf{\hat e}_n
    \end{pmatrix}
    \label{eq:ecliptic_to_parallel}
\end{equation}
By construction, at time $t_0$ we have
$\mathbf{\dot u}_0\,\bot\,\dot{\mathbf{u}}_0$ and the two components of
$\mathbf{u}(t)$ are then
\begin{align}
    u_\bot(t)      & \equiv \mathbf{u}(t)\cdot \mathbf{\hat e}_\bot= u_0 +
    \pi_E\,\delta\boldsymbol \zeta(t)\cdot\mathbf{\hat e}_\bot                                                                                                                                       \\
    u_\parallel(t) & \equiv \mathbf{u}(t)\cdot \mathbf{\hat e}_\parallel= (t-t_0)\,\dot{\mathbf{u}}_0\cdot\mathbf{\hat e}_\parallel+ \pi_E\,\delta\boldsymbol \zeta(t)\cdot\mathbf{\hat e}_\parallel
\end{align}
using the definitions of the unit vectors and
$\delta\boldsymbol \zeta(t)=
    \delta \zeta_e(t)\,\mathbf{\hat e}_e+\delta \zeta_n(t)\,\mathbf{\hat e}_n$, we have
\begin{align}
    u_\bot(t)      & = u_0 + \pi_E\,\cos\psi\,\delta \zeta_e(t) - \pi_E\,\sin\psi\,\delta \zeta_n(t)
    \label{eq:u_t_parallel1}                                                                         \\
    u_\parallel(t) & =(t-t_0)/t_E' + \pi_E\,\sin\psi\,\delta \zeta_e(t) +
    \pi_E\,\cos\psi\,\delta \zeta_n(t) \label{eq:u_t_parallel2}
\end{align}
where $t_E' =|\dot{\mathbf{u}}_0|=|\boldsymbol\mu_{LS}/\theta_E  + \pi_E\,\dot{\boldsymbol \zeta}(t_0')|$.
The model parameters which determine the magnification $A(t)$ are then
$\left(u_0,t_0,t'_E,\pi_E,\psi\right)$.
An alternative parametrization which is more commonly used is obtained by
defining components of a "microlensing parallax vector" as
\begin{align}
    \pi_{E,N} & \equiv \pi_E\cos\psi \\
    \pi_{E,E} & \equiv \pi_E\sin\psi
\end{align}
in which case we have
\begin{align}
    u_\bot(t)      & = u_0 + \pi_{E,N}\,\delta \zeta_e(t) - \pi_{E,E}\,\delta \zeta_n(t) \\
    u_\parallel(t) & =(t-t_0)\omega_E + \pi_{E,E}\,\delta\zeta_e(t) +
    \pi_{E,N}\,\delta\zeta_n(t)
\end{align}
and the model parameters are $\left(u_0,t_0,t_E',\pi_{E,E},\pi_{E,N}\right)$.

For reference, we can also write down the components of $\mathbf{u}(t)$ in
equatorial coordinates by inverting the rotation matrix in
Equation~\ref{eq:ecliptic_to_parallel} and applying it to
Equations~\ref{eq:u_t_parallel1} and \ref{eq:u_t_parallel2}, the two components
are then
\begin{align}
    u_e & =u_0\cos\psi + (t-t_0)\omega_E\sin\psi + \pi_E\,\delta\zeta_e(t)  \\
    u_n & =-u_0\sin\psi + (t-t_0)\omega_E\cos\psi + \pi_E\,\delta\zeta_n(t)
\end{align}

@@

\section{Model validation and comparison}
\label{sec:model_comparison}
\subsection{Cross validation}
% discuss vanilla k-fold CV, stress out just how important and central CV is to science
\subsubsection{Pareto smoothed importance sampling LOO}
\subsection{Model comparison}
% don't do it!

>>>>

% Define document class
\documentclass[12pt]{report}
\usepackage{aas_macros}

@@

That the same effect could be used to detect the presence of planets orbiting
around the lens star was first theorized by \citet{1964PhRv..133..835L} who
wrote that \emph{``the primary effect of planetary deflectors bound to stars
    other than the Sun will be to slightly perturb the lens action of these
    stars''} although he was also sceptical about the possibility of detection,
saying that \emph{``associated pulses would be so weak and infrequent and of
    such fleeting duration – perhaps a few hours – as to defy detection''}.
Gravitational lensing as a method for discovering exoplanets really took off
with the work of Paczy\'nski
\citep{1986ApJ...304....1P,1986ApJ...301..503P,1991ApJ...374L..37M} who also
coined the term \emph{microlensing}, which refers to gravitational lensing in a
regime where the images of the background object cannot be resolved but one can
nevertheless measure its magnification as a function of time. Microlensing as a
method for detecting exoplanets has some unique aspects. First, it is a one-off
event which happens on a timescale of a few minutes up to several months
depending on the distances to the background star and the lensing star and the
mass of the lens. In addition to the fact that there's only one chance to
observe such an event for it to happen at all requires extremely precise
alignment between the lens and the background star and the chance of observing
a stellar microlensing event is about one-in-a-million (CITE) for a typical
star within the Milky Way. To observe a planetary signal is about an order of
magnitude less likely than that. Hence, obtaining a decent sample of planetary
events requires continuous monitoring on the order of $10^{8}$ stars at the
time which way microlensing observations that to focus on the densest region of
the Milky Way -- the galactic bulge. Figure~\ref{fig:crowded_field} a picture
from of such a dense stellar field. Finally, in the vast majority of cases we
do not detect any light from the lens itself, the collected photons are from a
background star completely unrelated and distant from the lensing star. This is
very different from other exoplanet discovery methods and it means that we only
obtain dynamical properties of planets such their masses and periods.

@@

From Figure~\ref{fig:lens_geometry} it follows that the relationship between
the observed and the actual location of the source is given by
\begin{equation}
    \beta=\theta-2 R_s \frac{D_S-D_L}{D_LD_S}
    \frac{1}{\theta}\quad,
    \label{eq:lens_equation}
\end{equation}
which is known as the \emph{lens equation} or sometimes also the \emph{ray-tracing}
equation. If the source and the lens are perfectly aligned with respect to the observer,
then $\beta=0$ and $\theta\equiv\theta_E$ where
\begin{equation}
    \theta_E= \sqrt{ \frac{4GM}{c^2} \left( \frac{1}{D_L} - \frac{1}{D_S} \right)}=
    \sqrt{\kappa M\pi_{LS}}
    \label{eq:angular_einstein_radius}
\end{equation}
where $\kappa \equiv \frac{4 G}{c^{2} \mathrm{au}} \simeq 8.1 \frac{\mathrm{mas}}{M_{\odot}}$
and
\begin{equation}
    \pi_{LS}\equiv\pi_L - \pi_S=\frac{1\mathrm{au}}{D_L}-\frac{1\mathrm{au}}{D_S}
\end{equation}
is the relative lens-source parallax.

@@

To be able to evaluate Equation~\ref{eq:magnification_point_lens} in practice
we have to parametrize the position of the source on the sky $u$ as a function
of time. To derive an expression for $u(t)$ we assume the the motion of the
observer, lens and the source is rectilinear (acceleration can be neglected).
We take $\boldsymbol{\theta}_S$ to be the angular position of the source star
on the plane of the sky and $\boldsymbol{\mu}_L$ to be its proper motion
vector. Likewise for the lens. We thus have
\begin{align}
    \boldsymbol{\theta}_S(t) & =\boldsymbol{\theta}_{S, 0}+
    \left(t-t_{0}\right) \boldsymbol{\mu}_S
    \label{eq:source_position}                              \\
    \boldsymbol{\theta}_L(t) & =\boldsymbol{\theta}_{L, 0}
    +\left(t-t_{0}\right) \boldsymbol{\mu}_L\quad,
    \label{eq:lens_position}
\end{align}
where $t_0$ is some reference time.
The relative position vector of the lens with respect to the source is then
\begin{equation}
    \boldsymbol{u}(t) \equiv \frac{\boldsymbol{\theta}_L(t)
        -\boldsymbol{\theta}_S(t)}{\theta_E}=
    \frac{\boldsymbol{\theta}_{LS, 0}}{\theta_E}
    +\frac{t-t_{0}}{\theta_E} \boldsymbol{\mu}_{LS}
    \label{eq:relative_trajectory_no_parallax}
\end{equation}
where
$\boldsymbol{\theta}_{L S, 0}\equiv\boldsymbol{\theta}_{L,0}-\boldsymbol{\theta}_{S, 0}$
is the relative position at $t_0$ and
$\boldsymbol{\mu}_{LS}\equiv \boldsymbol{\mu}_{L}- \boldsymbol{\mu}_{S}$ is the
relative proper motion.
Since $\boldsymbol{\mu}_{LS}$ and $\boldsymbol{\theta}_{L S, 0}$ are
perpendicular to each other, it follows that the magnitude of the relative
separation is
\begin{equation}
    u(t)=\sqrt{u_0^2+ \left(\frac{t-t_0}{t_E}\right)^2}
\end{equation}
where we have defined $u_0\equiv |\boldsymbol{\theta}_{L S, 0}|/\theta_E$
and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
The magnification is then a function of three parameters $(t_0, u_0, t_E)$
and the resulting curve as a function of time is often called the Paczy\'nski curve
\citep{1986ApJ...304....1P,1986ApJ...301..503P}.
The magnification as a function of time for different impact parameters $u_0$
is shown in Figure~\ref{fig:paczynski_curve}. Notice the very steep fall off in
magnification as as we move away from the Einstein ring ($A(u)\propto
    1+2u^{-4}$).
In the limits of $u_0\rightarrow 0$ and $u_0\gg 1$  there is a continuous
mathematical degeneracy between the parameters $t_0$, $u_0$ and $t_E$
\citep{1997ApJ...487...55W}.

@@

\subsection{Observed flux}
The magnification $A(t)$ is not a direct observable in microlensing events. We
can only measure the flux of the source star which is usually contaminated with
the flux of other nearby stars in the dense stellar field and potentially also
with light from the lens and possible companions to the lens. The observed flux
is thus given by
\begin{equation}
    F(t)=F_S(t)+F_B(t)
    \label{eq:flux_observed}
\end{equation}
where $F_s$ is the flux of the source star and $F_B$ is the contamination flux also
called the \emph{blending flux}. To quantify the strength of the blending e can
define the emph{source flux fraction} $b_S$ as the fraction of total (unmagnified)
flux that is coming from the source star:
\begin{equation}
    b_S\equiv F_S/(F_S + F_B)
\end{equation}
For highly blended events $b_S\approx 0$ and in absence of blending $b_S=1$.
The source flux $F_S$ and the blending flux $F_B$ are generally highly correlated
and it is often preferable to use a slightly different parametrization proposed by
\citet{2009MNRAS.393..816D}. In this parametrization we use the
\emph{baseline flux} $F_\mathrm{base}$ and the difference between
the baseline flux and the flux at peak magnification $\Delta F\equiv F_S[A(u_0)-1]$
as the new parameters.
Equation~\ref{eq:flux_observed} then takes the form
\begin{equation}
    f=\Delta F\frac{A(t) - 1}{A(t_0)-1}+F_\mathrm{base}
    \label{eq:flux_dominik}
\end{equation}
It is generally straightforward to measure the flux difference $\Delta F$, the
baseline flux $F_\mathrm{base}$ and the center of the curve $t_0$ but getting a
good estimate of $t_E$ even outside of the regime $u_0\rightarrow 0$ and
$u_0\gg 1$ requires good photometric sampling along the ``wings'' of the light
curve \citep{2009MNRAS.393..816D}.

@@

\subsection{Annual parallax}
\label{ssec:single_lens_parallax}
In Equation~\ref{eq:relative_trajectory_no_parallax} we have assumed that the
relative motion of the lens with respect to the source on the plane of the sky
is rectilinear.
This is a reasonable approximation in a barycentric frame of
reference if the acceleration of the source and the lens can be neglected.
However, most microlensing events are observed from a non-inertial geocentric
frame (Earth) and sometimes also from multiple locations simultaneously. In those
cases we have to take into account parallax effects.
We differentiate between two kinds of parallax. First is the \emph{annual
    parallax} (sometimes also called the orbital parallax) which modifies the
lens-source relative motion vector $\boldsymbol\mu_{LS}$ due to the local
acceleration of Earth in its orbit. It is important for long timescale events
when the event timescale is equal to some substantial fraction of a year. The
effect of annual parallax is usually a slight modification of the shape of the
Paczy\'nski curve.

The second kind of parallax is the \emph{satellite parallax} which refers to a
difference in viewpoint when conducting simultaneous observations of the source
star from different locations separated by a substantial fraction of the
Einstein ring projected onto the observer plane $\tilde{r}_E\equiv
    D_{LS}\theta_E$ (where $D_{LS}^{-1}=D_L^{-1} - D_S^{-1}$). In practice this
means observing a microlensing event from Earth and a space based telescope
simultaneously. It was first proposed by \citet{1966MNRAS.134..315R}. There is
also the ``terrestrial parallax'' which is identical to the satellite parallax
except that it involves only ground-based observations separated by a large
fraction of Earth's diameter. A measurement of these parallax effects allows
for a partial breaking of the degeneracy in the event timescale $t_E$ by
providing a relationship between the mass and the distance to the lens
(assuming one can estimate the distance to the source star).

In this section I will describe the annual parallax effect. We start by
modifying Equations~\ref{eq:source_position} and \ref{eq:lens_position} to
include the projected motion of the Sun relative to the Earth on the plane of
the sky. We project $\mathbf s$ -- the position vector of the Sun relative to
Earth, onto a plane perpendicular to the line of sight towards the source star
(plane of the sky) which is defined by the unit vector $\boldsymbol{\hat n}$
normal to the plane. The unit vector $\boldsymbol{\hat n}$ depends on the sky
coordinates of the source star $(\alpha,\delta)$ (right ascension and
declination). We choose to work in geocentric equatorial coordinates defined by
the spherical unit vectors $\hat{\mathbf e}_n$, pointing North, and
$\hat{\mathbf e}_e$ pointing eastward such that the coordinate system is
right-handed. The unit vectors are defined by
\begin{align}
    \hat{\mathbf e}_e & = \hat{\mathbf z}\times \hat{\mathbf n}  \\
    \hat{\mathbf e}_n & = \hat{\mathbf n}\times\hat{\mathbf e}_e
\end{align}
The two components of $\mathbf s$ projected onto the plane of the sky are then
\begin{align}
    \zeta_e(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_e \\
    \zeta_n(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_n
\end{align}
In practice, we can retrieve $\mathbf{s}$ using NASA's JPL Horizons system and compute
the projected separation at any time $t$.
The angular positions of the source and the lens
(Equations~\ref{eq:source_position} and \ref{eq:lens_position}) then become
\begin{align}
    \boldsymbol{\theta}_S(t) & = \boldsymbol{\theta}_{S,0}+(t-t_0)\boldsymbol{\mu}_S
    +\pi_S\,\boldsymbol{\zeta}(t)                                                    \\
    \boldsymbol{\theta}_L(t) & = \boldsymbol{\theta}_{L,0}+(t-t_0)\boldsymbol{\mu}_L
    +\pi_L\,\boldsymbol{\zeta}(t)
\end{align}
where  $\pi_S\equiv 1\,\mathrm{au}/D_S$ is the source parallax, and $\pi_L$ is the
lens parallax.
The relative separation is then
\begin{equation}
    \boldsymbol{u}(t)= \frac{\boldsymbol\theta_{LS, 0}}{\theta_E}
    +\frac{t-t_0}{\theta_E}\boldsymbol{\mu}_{LS}+\pi_{E}\,\boldsymbol{\zeta}(t)
    \label{eq:relative_separation_parallax}
\end{equation}
where
\begin{equation}
    \pi_E\equiv \frac{\pi_{LS}}{\theta_E}
    \label{eq:pi_E}
\end{equation}

Since annual parallax is a higher order effect affecting the apparent
trajectory of the source on the sky, it makes sense to decompose the trajectory
as a sum of rectilinear motion plus a deviation due to parallax. Following
\citet{2004ApJ...606..319G} we define the position offset of the Sun on the
plane of the sky relative to its position at a particular time $t_0'$ as

@@

\begin{figure}
    \centering
    \includegraphics[width=0.3\linewidth]{static/microlensing/microlensing_parallax_coordinates.pdf}
    \caption{A coordinate system
        $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ parallel to the source trajectory at time $t_0$. The orange curve
        represents the source trajectory relative to the lens at the origin. The
        coordinate system $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ is related to the equatorial coordinates $(\mathbf{\hat
                e}_e,\mathbf{\hat e}_n)$ by a rotation through an angle $\psi$.}
    \label{fig:parallax2}
\end{figure}

To evaluate Equation~\ref{eq:relative_separation_parallax_decomposed}, we need
to choose a suitable basis. A natural coordinate system for describing the
trajectory of the source relative to the lens is one defined by unit vectors
$(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ where $\mathbf{\hat e}_\parallel$ is parallel to the trajectory $\mathbf{u}(t)$ at time $t_0$
(Figure~\ref{fig:parallax2}). We define the unit vectors as
\begin{equation}
    \mathbf{\hat e}_\bot\equiv \frac{\mathbf{u}_0}{|\mathbf{u}_0|}\quad,
    \mathbf{\hat e}_\parallel\equiv \frac{\mathbf{\hat n}\times\mathbf{u}_0}{|\mathbf{u}_0|}\quad
\end{equation}
and the coordinate system $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ is related to ecliptic coordinates via a simple rotation through an
angle $\psi$
\begin{equation}
    \begin{pmatrix}
        \mathbf{\hat e}_\bot \\
        \mathbf{\hat e}_\parallel
    \end{pmatrix}
    =
    \begin{pmatrix}
        \cos\psi & -\sin\psi \\
        \sin\psi & \cos\psi
    \end{pmatrix}
    \begin{pmatrix}
        \mathbf{\hat e}_e \\
        \mathbf{\hat e}_n
    \end{pmatrix}
    \label{eq:ecliptic_to_parallel}
\end{equation}
By construction, at time $t_0$ we have
$\mathbf{u}_0\,\bot\,\dot{\mathbf{u}}_0$ and the two components of
$\mathbf{u}(t)$ are then
\begin{align}
    u_\bot(t)      & \equiv \mathbf{u}(t)\cdot \mathbf{\hat e}_\bot= u_0 +
    \pi_E\,\delta\boldsymbol \zeta(t)\cdot\mathbf{\hat e}_\bot                                                                                                                                       \\
    u_\parallel(t) & \equiv \mathbf{u}(t)\cdot \mathbf{\hat e}_\parallel= (t-t_0)\,\dot{\mathbf{u}}_0\cdot\mathbf{\hat e}_\parallel+ \pi_E\,\delta\boldsymbol \zeta(t)\cdot\mathbf{\hat e}_\parallel
\end{align}
using the definitions of the unit vectors and
$\delta\boldsymbol \zeta(t)=
    \delta \zeta_e(t)\,\mathbf{\hat e}_e+\delta \zeta_n(t)\,\mathbf{\hat e}_n$, we have
\begin{align}
    u_\bot(t)      & = u_0 + \pi_E\,\cos\psi\,\delta \zeta_e(t) - \pi_E\,\sin\psi\,\delta \zeta_n(t)
    \label{eq:u_t_parallel1}                                                                         \\
    u_\parallel(t) & =(t-t_0)/t_E' + \pi_E\,\sin\psi\,\delta \zeta_e(t) +
    \pi_E\,\cos\psi\,\delta \zeta_n(t) \label{eq:u_t_parallel2}
\end{align}
where $t_E' =|\dot{\mathbf{u}}_0|=|\boldsymbol\mu_{LS}/\theta_E  + \pi_E\,\dot{\boldsymbol \zeta}(t_0')|$.
The model parameters which determine the magnification $A(t)$ are then
$\left(u_0,t_0,t'_E,\pi_E,\psi\right)$. Notice that $u_0$ in this case can also be negative.

An alternative parametrization which is more commonly used is obtained by
defining components of a "microlensing parallax vector" as
\begin{align}
    \pi_{E,N} & \equiv \pi_E\cos\psi \\
    \pi_{E,E} & \equiv \pi_E\sin\psi
\end{align}
in which case we have
\begin{align}
    u_\bot(t)      & = u_0 + \pi_{E,N}\,\delta \zeta_e(t) - \pi_{E,E}\,\delta \zeta_n(t) \\
    u_\parallel(t) & =(t-t_0)/t_E' + \pi_{E,E}\,\delta\zeta_e(t) +
    \pi_{E,N}\,\delta\zeta_n(t)
\end{align}
and the model parameters are $\left(u_0,t_0,t_E',\pi_{E,E},\pi_{E,N}\right)$.

For reference, we can also write down the components of $\mathbf{u}(t)$ in
equatorial coordinates by inverting the rotation matrix in
Equation~\ref{eq:ecliptic_to_parallel} and applying it to
Equations~\ref{eq:u_t_parallel1} and \ref{eq:u_t_parallel2}, the two components
are then
\begin{align}
    u_e & =u_0\cos\psi + (t-t_0)/t_E'\sin\psi + \pi_E\,\delta\zeta_e(t)  \\
    u_n & =-u_0\sin\psi + (t-t_0)/t_E'\cos\psi + \pi_E\,\delta\zeta_n(t)
\end{align}

@@

\section{Model validation and comparison}
\label{sec:model_comparison}

``Furthermore, in the future, a blind search for planetary signals will not
turn up all potential competing models unless those models
are specifically considered.''\citep{2017AJ....153..129J}.
\citet{2021AJ....162...59R}
\citet{2020AJ....160...17H}

\subsection{Cross validation}
% discuss vanilla k-fold CV, stress out just how important and central CV is to science
\subsubsection{Pareto smoothed importance sampling LOO}
\subsection{Model comparison}
% don't do it!
